---
title: "AsMi_BayesLeaves_Thyme_Laurel"
author: "Michelle DePrenger-Levin"
date: "April 12, 2018"
output: html_document
---


```{r}
library(R2jags)
require(lme4)
#install.packages('Rcpp')
library(Rcpp)
library(AICcmodavg)
library(lattice)
HDIofMCMC = function( sampleVec , credMass=0.95 ) {
  # Computes highest density interval from a sample of representative values,
  #   estimated as shortest credible interval.
  # Arguments:
  #   sampleVec
  #     is a vector of representative values from a probability distribution.
  #   credMass
  #     is a scalar between 0 and 1, indicating the mass within the credible
  #     interval that is to be estimated.
  # Value:
  #   HDIlim is a vector containing the limits of the HDI
  sortedPts = sort( sampleVec )
  ciIdxInc = ceiling( credMass * length( sortedPts ) )
  nCIs = length( sortedPts ) - ciIdxInc
  ciWidth = rep( 0 , nCIs )
  for ( i in 1:nCIs ) {
    ciWidth[ i ] = sortedPts[ i + ciIdxInc ] - sortedPts[ i ]
  }
  HDImin = sortedPts[ which.min( ciWidth ) ]
  HDImax = sortedPts[ which.min( ciWidth ) + ciIdxInc ]
  HDIlim = c( HDImin , HDImax )
  return( HDIlim )
}
plotPost = function( paramSampleVec , cenTend=c("mode","median","mean")[1] , 
                     compVal=NULL, ROPE=NULL, credMass=0.95, HDItextPlace=0.7, 
                     xlab=NULL , xlim=NULL , yaxt=NULL , ylab=NULL , 
                     main=NULL , cex=NULL , cex.lab=NULL ,
                     col=NULL , border=NULL , showCurve=FALSE , breaks=NULL , 
                     ... ) {
  # Override defaults of hist function, if not specified by user:
  # (additional arguments "..." are passed to the hist function)
  if ( is.null(xlab) ) xlab="Param. Val."
  if ( is.null(cex.lab) ) cex.lab=1.5
  if ( is.null(cex) ) cex=1.4
  if ( is.null(xlim) ) xlim=range( c( compVal , ROPE , paramSampleVec ) )
  if ( is.null(main) ) main=""
  if ( is.null(yaxt) ) yaxt="n"
  if ( is.null(ylab) ) ylab=""
  if ( is.null(col) ) col="skyblue"
  if ( is.null(border) ) border="white"
  
  # convert coda object to matrix:
  if ( class(paramSampleVec) == "mcmc.list" ) {
    paramSampleVec = as.matrix(paramSampleVec)
  }
  
  summaryColNames = c("ESS","mean","median","mode",
                      "hdiMass","hdiLow","hdiHigh",
                      "compVal","pGtCompVal",
                      "ROPElow","ROPEhigh","pLtROPE","pInROPE","pGtROPE")
  postSummary = matrix( NA , nrow=1 , ncol=length(summaryColNames) , 
                        dimnames=list( c( xlab ) , summaryColNames ) )
  
  # require(coda) # for effectiveSize function
  postSummary[,"ESS"] = effectiveSize(paramSampleVec)
  
  postSummary[,"mean"] = mean(paramSampleVec)
  postSummary[,"median"] = median(paramSampleVec)
  mcmcDensity = density(paramSampleVec)
  postSummary[,"mode"] = mcmcDensity$x[which.max(mcmcDensity$y)]
  
  HDI = HDIofMCMC( paramSampleVec , credMass )
  postSummary[,"hdiMass"]=credMass
  postSummary[,"hdiLow"]=HDI[1]
  postSummary[,"hdiHigh"]=HDI[2]
  
  # Plot histogram.
  cvCol = "darkgreen"
  ropeCol = "darkred"
  if ( is.null(breaks) ) {
    if ( max(paramSampleVec) > min(paramSampleVec) ) {
      breaks = c( seq( from=min(paramSampleVec) , to=max(paramSampleVec) ,
                       by=(HDI[2]-HDI[1])/18 ) , max(paramSampleVec) )
    } else {
      breaks=c(min(paramSampleVec)-1.0E-6,max(paramSampleVec)+1.0E-6)
      border="skyblue"
    }
  }
  if ( !showCurve ) {
    par(xpd=NA)
    histinfo = hist( paramSampleVec , xlab=xlab , yaxt=yaxt , ylab=ylab ,
                     freq=F , border=border , col=col ,
                     xlim=xlim , main=main , cex=cex , cex.lab=cex.lab ,
                     breaks=breaks , ... )
  }
  if ( showCurve ) {
    par(xpd=NA)
    histinfo = hist( paramSampleVec , plot=F )
    densCurve = density( paramSampleVec , adjust=2 )
    plot( densCurve$x , densCurve$y , type="l" , lwd=5 , col=col , bty="n" ,
          xlim=xlim , xlab=xlab , yaxt=yaxt , ylab=ylab ,
          main=main , cex=cex , cex.lab=cex.lab , ... )
  }
  cenTendHt = 0.9*max(histinfo$density)
  cvHt = 0.7*max(histinfo$density)
  ROPEtextHt = 0.55*max(histinfo$density)
  # Display central tendency:
  mn = mean(paramSampleVec)
  med = median(paramSampleVec)
  mcmcDensity = density(paramSampleVec)
  mo = mcmcDensity$x[which.max(mcmcDensity$y)]
  if ( cenTend=="mode" ){ 
    text( mo , cenTendHt ,
          bquote(mode==.(signif(mo,3))) , adj=c(.5,0) , cex=cex )
  }
  if ( cenTend=="median" ){ 
    text( med , cenTendHt ,
          bquote(median==.(signif(med,3))) , adj=c(.5,0) , cex=cex , col=cvCol )
  }
  if ( cenTend=="mean" ){ 
    text( mn , cenTendHt ,
          bquote(mean==.(signif(mn,3))) , adj=c(.5,0) , cex=cex )
  }
  # Display the comparison value.
  if ( !is.null( compVal ) ) {
    pGtCompVal = sum( paramSampleVec > compVal ) / length( paramSampleVec ) 
    pLtCompVal = 1 - pGtCompVal
    lines( c(compVal,compVal) , c(0.96*cvHt,0) , 
           lty="dotted" , lwd=2 , col=cvCol )
    text( compVal , cvHt ,
          bquote( .(round(100*pLtCompVal,1)) * "% < " *
                    .(signif(compVal,3)) * " < " * 
                    .(round(100*pGtCompVal,1)) * "%" ) ,
          adj=c(pLtCompVal,0) , cex=0.8*cex , col=cvCol )
    postSummary[,"compVal"] = compVal
    postSummary[,"pGtCompVal"] = pGtCompVal
  }
  # Display the ROPE.
  if ( !is.null( ROPE ) ) {
    pInROPE = ( sum( paramSampleVec > ROPE[1] & paramSampleVec < ROPE[2] )
                / length( paramSampleVec ) )
    pGtROPE = ( sum( paramSampleVec >= ROPE[2] ) / length( paramSampleVec ) )
    pLtROPE = ( sum( paramSampleVec <= ROPE[1] ) / length( paramSampleVec ) )
    lines( c(ROPE[1],ROPE[1]) , c(0.96*ROPEtextHt,0) , lty="dotted" , lwd=2 ,
           col=ropeCol )
    lines( c(ROPE[2],ROPE[2]) , c(0.96*ROPEtextHt,0) , lty="dotted" , lwd=2 ,
           col=ropeCol)
    text( mean(ROPE) , ROPEtextHt ,
          bquote( .(round(100*pLtROPE,1)) * "% < " * .(ROPE[1]) * " < " * 
                    .(round(100*pInROPE,1)) * "% < " * .(ROPE[2]) * " < " * 
                    .(round(100*pGtROPE,1)) * "%" ) ,
          adj=c(pLtROPE+.5*pInROPE,0) , cex=1 , col=ropeCol )
    
    postSummary[,"ROPElow"]=ROPE[1] 
    postSummary[,"ROPEhigh"]=ROPE[2] 
    postSummary[,"pLtROPE"]=pLtROPE
    postSummary[,"pInROPE"]=pInROPE
    postSummary[,"pGtROPE"]=pGtROPE
  }
  # Display the HDI.
  lines( HDI , c(0,0) , lwd=4 , lend=1 )
  text( mean(HDI) , 0 , bquote(.(100*credMass) * "% HDI" ) ,
        adj=c(.5,-1.7) , cex=cex )
  text( HDI[1] , 0 , bquote(.(signif(HDI[1],3))) ,
        adj=c(HDItextPlace,-0.5) , cex=cex )
  text( HDI[2] , 0 , bquote(.(signif(HDI[2],3))) ,
        adj=c(1.0-HDItextPlace,-0.5) , cex=cex )
  par(xpd=F)
  #
  return( postSummary )
}

logit2prob <- function(logit){
  odds <- exp(logit)
  prob <- odds / (1 + odds)
  return(prob)
}

```


#Set up long-form data   
variable indiviuals (1 or 0), per variable plots (g) per site (Q) and variable years (t) per plot.     
```{r}
#Climate at the Q level, covariables at the i level. 
siteN <- c(6,5,5,4,5,5)
B0 <- 3 # grand mean
 
sigmai <- 0.05 # variance among individuals
sigmaQ <- .3 # variance in mean among sites, spatial differences 
sigmag <- .1 # variance around mean associated with plots, more similar 
sigmaf <- .05 # variance around plots with factor or without

B1_plot <- c(seq(-.5,.5,length.out = 30))
B1_site <- c(seq(-1,1,length.out = 6))
B_environ <- c(0.7,5)
B_w <- c(3,-0.5)
B_cofactor <- c(-3,0.1)
```

Not beta! doesn't need to be [0-1]    

      B1 <- rbeta(6, (B0^2 - B0^3 - B0*sigmaQ^2)/sigmaQ^2,     # true values of Site means
                  (B0-2*B0^2+B0^3-sigmaQ^2+B0*sigmaQ^2)/sigmaQ^2) #between 0 and 1
      B2 <- do.call(c,lapply(1:6, function(b){
        rbeta(siteN[b],(B1[b]^2 - B1[b]^3 - B1[b]*sigmag^2)/sigmag^2, # true values of plot means
              (B1[b]-2*B1[b]^2+B1[b]^3-sigmag^2+B1[b]*sigmag^2)/sigmag^2)})) # with true variance around each site

# plot factors
```{r}
B0_plotfence <- 0.25 #variance around mean for plots with vs. without fencing (or burn treatment), offset for fenced
```

# individual measurements, w
```{r}
B4 <- 0.3 # slope for length increases survival?
B5 <- 0.55 # fruit increases survival
```

# site measurements, x
```{r}
B6 <- 0.4 #more moisture means more survival
B7 <- 0.9 #more moisure, bigger slope change than the first moisture one
B8 <- -0.5 #hotter bad
```

#Data structure to match different lengths of indexing   
```{r}
Q <- sort(sample(c(1,2,5,15,19,26), size=1000, replace=TRUE)) # sort to allow random number of appropraite plot repeats (individuals: i) for the right plots (g) per site (Q)
#Variable number of plots per site, 30 total plots
g <- c(sample(1:6,size=table(Q)[1], replace=TRUE),
       sample(c(1:5)+6,size=table(Q)[2], replace=TRUE),
       sample(c(1:5)+11,size=table(Q)[3], replace=TRUE),
       sample(c(1:4)+16,size=table(Q)[4], replace=TRUE),
       sample(c(1:5)+20,size=table(Q)[5], replace=TRUE),
       sample(c(1:5)+25,size=table(Q)[6], replace=TRUE))
t <- do.call(c,lapply(1:30, function(plot){
  if(plot<15){
    sample(2000:2007,size=table(g)[plot], replace=TRUE)
  } else {
    sample(1999:2008,size=table(g)[plot], replace=TRUE)
  }}) ) #subset of years, different lengths of years

#Individual measurements [i]
w1 <- runif(1000, 0, 90) #length of longest stem, or zero when y is 0
w2 <- rnorm(1000,200,50) # number of fruit per individual, zero when length < 15, multiplied by survival vector will make 0 have 0 length and 1 have simluated random length

#create climate varaibles, Site measurements [Q]
gQt <- data.frame(Q,g,t)[with(data.frame(Q,g,t), order(Q,g,t)),]
climate.sim <- data.frame(unique(gQt), x1 = rnorm(nrow(unique(gQt)),100,5), #precip december
                          x2 = rnorm(nrow(unique(gQt)), 80,2),  #precip November
                          x3 = rnorm(nrow(unique(gQt)), 30, 3)) #max temp July

#plot level factor fenced or not (or burn treatment) [g]
G1 <- data.frame(g= unique(g),w.g1 = rbinom(30,1,0.5))

data.df <- data.frame(g,Q,t,w1,w2)
data.df <- merge(data.df,climate.sim,by=c("g","Q","t")) #merge with climate per site per year
data.df <- merge(data.df, G1, by="g") # merge with plot level factor (y/n fencing)

#How to simulate individual covariates based on simulated survival data??
data.df$w2[data.df$w1 < 15] <- 0 #can't flower if smaller than 15cm
hist(data.df$w1)

data.df <- data.df[with(data.df, order(Q,g,t)),] 

nsite <- length(table(data.df$Q)) #number of site
nplot <- aggregate(g~Q,data=data.df, function(x) length(unique(x)))$g #plots per site
nyear <- aggregate(t~g,data=data.df, function(x) length(unique(x)))$t #years per plot
#nobs <- aggregate(g~t+Q, data=data.df, function(x) length(x))
table(data.df$g,data.df$Q)
sum(table(data.df$g,data.df$Q))
```

###same number per each
# 5 plots per site
# 34 indivudals per plot
# 1020 indiviudals total, 34 per plot


y_site <- unlist(rep(lapply(1:nsite, function(Q) rnorm(1,B0 + B1_site[Q],sigmaQ)), each=5)) #
y_plot <- unlist(rep(lapply(1:30, function(g){rnorm(1, y_site[g]+B1_plot[g], sigmag)}), each=34))
y_plot_prob <- logit2prob(y_plot)
y <- unlist(lapply(1:1020, function(i){ rbinom(1,1,logit2prob(y_plot[i]))})) 






# f(survival) ~ grand mean from + site means from  + plot means + difference in treatment of plots +
#               w covariates of individuals + x covariates of sites
mu <- B0 + (B1*data.df$g) + (B2*data.df$Q) + (B3*data.df$w.g1) + 
  B4*data.df$w1 + B4*data.df$w2 +   #w vector  
  B5*data.df$x1 + B6*data.df$x2 + B7*data.df$x3  #x vector
#treatment of fence changes snow pack so might interact with slope of precip

y <- do.call(c,lapply(1:30, function(x) {
  rbinom(length(data.df$y[data.df$g==x]),1,1-(sigmag^2/(B0+B2[x])))
}))

plot(mu, y)
plot(y~g, data = data.df)
oddstest <- data.frame(y.1 = y,mu = mu, data.df)

survival <- tapply(oddstest$y.1,oddstest$Q,function(x) log((sum(x)/length(x))/
                                                         (1-(sum(x)/length(x)))) )
#Odds ratio
plot(survival)

str(data.df)




#Defining the model
prevalence <- function(){
  # Likelihood:
  for(k in 1:nsite){
    for(i in 1:nplot[k]){ # Within each site k, the number of plots is denoted by the vector nplot[k]
      for(t in 1:nyear[k,i]){ # Within each plot i of site k, the number years it is surveyed is denoted by nyear[k,i]
        for(j in 1:nobs[k,i,t]){ # The number of observations at plot i in year t is denoted by nobs[k,i,t]
          y[k,i,j,t] ~ dbern(1-(1-pi[k,i,t])^n[k,i,j,t])
        }
        pi[k,i,t] <- ilogit(beta0[k,i] + beta1*x1[k,i,t] + beta2*x2[k,i,t])
      }
      beta0[k,i] ~ dnorm(alpha[k], 1/sigma[k]^2)
    }
    alpha[k] ~ dnorm(mu, 1/sigma0^2)
  }
  
  # Priors:
  beta1 ~ dnorm(0, .01)
  beta2 ~ dnorm(0, .01)
  mu ~ dnorm(0, .01)
  sigma0 ~ dunif(0,10)
  for(k in 1:nsite){
    sigma[k] ~ dunif(0,10)
  }
}

```















#Actual data
```{r}
load("P:/My Documents/BDA_Spring2018/AsMi_subset.Rda")
asmi.climate<-read.csv("P:/hackathon/GitPVA/datasets/asmi.c.csv")
#load("AsMi_subset.Rda") #asmi.toshare
head(asmi.toshare)
```

$\beta_0$    
y^G^~i,Q,t+1~ 

#Climate variable importance  
```{r}
#asmi.climate <- read.csv("asmi.c.csv")
head(asmi.climate)

asmi.merge <- merge(asmi.toshare, asmi.climate, by.x = "AsMi_plot_id", by.y = "Plot")
head(asmi.merge)

#test variable importance 
```


#StagePVA    
I'd like to change this from having dormancy as a stage (and then allowing some death to dormants) to detectability
```{r}
StagePVA <- function(df,dormancy = 1){
  
  years <- sort(unique(df$year)) #start years of each transition
  
  # Create a list to hold output 
  names <- c("plot.matrix",   #Stage PVA per plot
             "pro.matrix","fenced.pro.matirx","notfenced.pro.matrix",    #Stage PVA overall
             "Site", "fenced", "notfenced") #Stage PVA per site: fenced vs. not fenced plots split per site
  SiteMatrix <- vector("list", length(names))
  names(SiteMatrix) <- names
  
  require(popbio)
  
  # Individuals that are new at the current census are considered seedlings, some are reproductive
  # These measure the fertility (fruit output) of seedlings
  # "Seedlings" that are reproductive came from seed at an earlier time, won't be reproductive 
  # if they were a seedling that year - should add to seed stage somehow....
  
  ## 1
  # overall
  seedlingsfert <- c()
  fencedseedlingsfert <- c()
  notfencedseedlingsfert <- c()
  
  # projection matrices of of class specific vital rates
  pro.matrix <- vector("list", length(years))
  names(pro.matrix) <- years
  fenced.pro.matrix <- vector("list", length(years))
  names(fenced.pro.matrix) <- years
  notfenced.pro.matrix <- vector("list", length(years))
  names(notfenced.pro.matrix) <- years
  
  ## Fencing
  for(i in years){
    
    # Not split by site, overall
    fert <- subset(df, x == i)
    fencedfert <- subset(df, x == i & df$fenced == "y")
    notfencedfert <- subset(df, x == i & df$fenced == "n")		
    
    # Pre-breeding census
    # ?projection.matrix uses a column with a stage name as a fertility measure per plant 
    # fruit production per individual as a percent of the total production that year. Time t
    # this times the number of seedlings that survived the next year
    
    seedlings <- nrow(subset(df, x == i+1 & df$stage== "seedling"))
    fencedseedlings <- nrow(subset(df, x == i+1 & df$stage== "seedling" & df$fenced == "y"))
    notfencedseedlings <- nrow(subset(df, x == i+1 & df$stage== "seedling" & df$fenced == "n"))
    
    # Adding a fertility column to seedlings. 
    # instead could Do something like if seedling and length > 10cm then call it vegetative/repro and add to 
    #  seed count???
    fert$seedling <- seedlings * (fert$fruits / sum(fert$fruits, na.rm = T))
    fencedfert$seedling <- fencedseedlings * (fencedfert$fruits / sum(fencedfert$fruits, na.rm = T))
    notfencedfert$seedling <- notfencedseedlings * (notfencedfert$fruits / sum(notfencedfert$fruits, na.rm = T))
    
    # Add some survival to dormant individuals other than 100% since we only note dormat indivdiuals when they are
    #seen above ground again so 100% survive # the column sums to 1, want some to transition to dead, 
    # lowering dormant to dormant
    death.to.dormants_fert <- projection.matrix(fert)[length(unique(df$stage)),length(unique(df$stage))]*dormancy
    death.to.dormants_fencedfert <- projection.matrix(fencedfert)[length(unique(df$stage)),length(unique(df$stage))]*dormancy
    death.to.dormants_notfencedfert <- projection.matrix(notfencedfert)[length(unique(df$stage)),length(unique(df$stage))]*dormancy
    
    
    pro.matrix[[as.character(i)]] <- projection.matrix(fert, add = c(length(unique(df$stage)),
                                                                     length(unique(df$stage)), 
                                                                     death.to.dormants_fert))
    fenced.pro.matrix[[as.character(i)]] <- projection.matrix(fencedfert, add = c(length(unique(df$stage)),
                                                                                  length(unique(df$stage)), 
                                                                                  death.to.dormants_fencedfert))
    notfenced.pro.matrix[[as.character(i)]] <- projection.matrix(notfencedfert, add = c(length(unique(df$stage)),
                                                                                        length(unique(df$stage)), 
                                                                                        death.to.dormants_notfencedfert))
  }
  
  SiteMatrix$pro.matrix <- pro.matrix
  SiteMatrix$fenced.pro.matrix <- fenced.pro.matrix
  SiteMatrix$notfenced.pro.matrix <- notfenced.pro.matrix
  
  
  ## 2
  # Projection matrices divdied by Site
  
  Site.matrix <- vector("list", length(years))
  fenced.Site.matrix <- vector("list", length(years))
  notfenced.Site.matrix <- vector("list", length(years))
  
  promatrix <- vector("list", length(years))
  names(promatrix) <- years
  fenced.promatrix <- vector("list", length(years))
  names(fenced.promatrix) <- years
  notfenced.promatrix <- vector("list", length(years))
  names(notfenced.promatrix) <- years
  
  #Set variables
  sites <- unique(df$site)
  
  ## Fencing by site
  for(j in sites){
    for(i in years){
      
      # df$site is Site
      # x is Year
      rm(fencedfert)
      fencedfert <- subset(df, x == i & df$site == j & df$fenced == "y")
      fencedseedlings <- nrow(subset(df, x == i+1 & df$stage == "seedling" & df$site == j & 
                                       df$fenced == "y"))
      
      # no fencing in site 15 creates errors. Fencing removed from all in 2015
      if(nrow(fencedfert) == 0){ 
        fencedfert$seedling <- 0 * (fencedfert$fruits / sum(fencedfert$fruits, na.rm = T))
        fencedfert$seedling[is.nan(fencedfert$seedling)] <- 0
        fenced.promatrix[[as.character(i)]] <- projection.matrix(fencedfert)
      } else {
        fencedfert$seedling <- fencedseedlings * (fencedfert$fruits / sum(fencedfert$fruits, 
                                                                          na.rm = T))
        fencedfert$seedling[is.nan(fencedfert$seedling)] <- 0
        fenced.promatrix[[as.character(i)]] <- projection.matrix(fencedfert)}	# end if else
      
      rm(fert)
      rm(notfencedfert)
      fert <- subset(df, x == i & df$site == j)
      notfencedfert <- subset(df, x == i & df$site == j & df$fenced == "n")
      
      # ?projection.matrix uses a column with a stage name as a fertility measure per plant 
      # fruit production per individual as a percent of the total production that year. Time t
      # this times the number of seedlings that survived the next year
      
      seedlings <- nrow(subset(df, x == i+1 & df$stage == "seedling" & df$site == j))
      notfencedseedlings <- nrow(subset(df, x == i+1 & df$stage == "seedling" & df$site == j &
                                          df$fenced == "n"))
      
      fert$seedling <- seedlings * (fert$fruits / sum(fert$fruits, na.rm = T))
      fert$seedling[is.nan(fert$seedling)] <- 0
      fert$seedling[is.na(fert$seedling)] <- 0
      notfencedfert$seedling <- notfencedseedlings * (notfencedfert$fruits / sum(notfencedfert$fruits,
                                                                                 na.rm = T))
      notfencedfert$seedling[is.nan(notfencedfert$seedling)] <- 0
      notfencedfert$seedling[is.na(notfencedfert$seedling)] <- 0		 
      
      death.to.dormants <- projection.matrix(fert)[length(unique(df$stage)),
                                                   length(unique(df$stage))]*dormancy
      
      promatrix[[as.character(i)]] <- projection.matrix(fert, 
                                                        add = c(length(unique(df$stage)),
                                                                length(unique(df$stage)), 
                                                                death.to.dormants))
      notfenced.promatrix[[as.character(i)]] <- projection.matrix(notfencedfert, 
                                                                  add = c(length(unique(df$stage)),
                                                                          length(unique(df$stage)), 
                                                                          death.to.dormants))
    }
    Site.matrix[[as.character(j)]] <- promatrix
    fenced.Site.matrix[[as.character(j)]] <- fenced.promatrix
    notfenced.Site.matrix[[as.character(j)]] <- notfenced.promatrix
  }
  
  # add each year list of matrices to each by site and for only fenced and only not fenced
  SiteMatrix$Site <- Site.matrix
  SiteMatrix$fenced <- fenced.Site.matrix
  SiteMatrix$notfenced <- notfenced.Site.matrix
  
  ## 3
  # Projection matrices divdied by Plot
  
  plot.matrix <- vector("list", length(years))
  names(plot.matrix) <- years
  plotpromatrix <- vector("list", length(years))
  names(plotpromatrix) <- years
  
  #Set variables
  Plots <- unique(df$plot)
  
  # sample size might be too small per plot
  ## Plot
  for(j in Plots){
    for(i in years){
      
      # df$year is Year
      fertplot <- subset(df, df$year == i & df$plot == j)
      
      # ?projection.matrix uses a column with a stage name as a fertility measure per plant 
      # fruit production per individual as a percent of the total production that year. Time t
      # this times the number of seedlings that survived the next year
      
      seedlings <- nrow(subset(df, df$year == i+1 & df$stage == "seedling" & df$plot == j))
      fert$seedling <- seedlings * (fert$fruits / sum(fert$fruits, na.rm = T))
      fert$seedling[is.nan(fert$seedling)] <- 0
      fert$seedling[is.na(fert$seedling)] <- 0
      
      death.to.dormants <- projection.matrix(fertplot)[length(unique(df$stage)),
                                                       length(unique(df$stage))]*dormancy 
      
      plotpromatrix[[as.character(i)]] <- projection.matrix(fert, add = c(length(unique(df$stage)),
                                                                          length(unique(df$stage)), 
                                                                          death.to.dormants))
      
    }
    
    plot.matrix[[as.character(j)]] <- plotpromatrix
  }
  
  # add each year list of matrices to each by site and for only fenced and only not fenced
  SiteMatrix$plot.matrix <- plot.matrix
  
  # The list returned from the function
  SiteMatrix
}

```



DAG #2   
```{r}
currentyr <- as.numeric(format(as.Date(Sys.Date(),format="%Y-%m-%d"), "%Y")) -1
rawdatapath <- paste("Q:/Research/All_Projects_by_Species/Astragalus SPECIES/Astragalus_microcymbus/Asmi_Excel/Yearly Summaries/", 
                     currentyr,"_asmi/RawData_",
                     currentyr, ".csv", collapse = '', sep = '')

asmi.raw <- read.csv(path.expand(rawdatapath), na.strings = "na")

# No longer adding climate data to the database
asmi.raw <- asmi.raw[,grep(paste(c("Temp","Rain","Snow","Aug.Jul"),collapse="|"), names(asmi.raw),
                           value = TRUE, invert = TRUE)]
asmi.raw <- asmi.raw[asmi.raw$status != "",]
asmi.raw$status <- factor(asmi.raw$status)
#reset factors of browsing to eliminate " "
asmi.raw$Browsing...Status[asmi.raw$Browsing...Status == "mammal"] <- "Mammal"
asmi.raw$Browsing...Status <- factor(asmi.raw$Browsing...Status)

#reset factors for fence
asmi.raw$fence <- factor(asmi.raw$fence)

```


```{r}
#library("reshape2")

head(asmi.climate)
#asmi.climate$PrevYearMonth <- paste(asmi.climate$Prev12,asmi.climate$Month, sep="_")
climate.wide <- reshape(asmi.climate[,-8], 
                        idvar= c("Plot","Site","Creek","Site.1","Prev12"), timevar="Month", 
                        direction="wide")

asmi.merge <- merge(asmi.raw, climate.wide, by.x = c("AsMi_plot_id","AsMi_site_id","year"), 
                    by.y = c("Plot","Site","Prev12"))

#for indexing, something like:
# or if that doesn't work, sort first and then these will be sequential numbers
plot610 <- which(asmi.merge$AsMi_plot_id==610) #over the length of that
asmi.merge[plot610,]

length.plots.indexing <- sapply(unique(asmi.merge$AsMi_plot_id), function(plot){
  index <- which(asmi.merge$AsMi_plot_id == plot)
  out <- c(min(index),max(index))
  out
})

asmi.merge[1:291,]

for(i in 1:length(plot610)){
  x[plot610[i]]
}

plots.g <- 30
sites.Q <- 6 # 5,5,6,4,5
years.t.g <- apply(table(asmi.merge$year,asmi.merge$AsMi_plot_id),2,function(x)length(x[x>0])) #years per plot
samplesize.n_g <- apply(table(asmi.merge$AsMi_tag_id,asmi.merge$AsMi_plot_id),2,
                      function(x) length(x[x>0])) #tags per plot

#do for each plot seperately
data.asmi1 <- asmi.merge[,c(1,2,3,6,8,46,49,52,55,58)]
# model with fall precip
jagsdata1 <- with(data.asmi1, list(y = length, x1 = fruit, x2 = Precip.8, x3 = Precip.9,
                                  x4 = Precip.10, x5 = Precip.11, x6 = Precip.12,
                                  N = length(length)))

mu <- predict(lm(length ~ fruit+Precip.8+Precip.9+Precip.10+Precip.11+Precip.12 ,data=data.asmi1))


lm1_jags <- function(){
  #Likelihood:
  for(i in 1:N){
    y[i] ~ dnorm(mu[i], tau)
    mu[i] <- beta[1] + x1[i]*beta[2] + x2[i]*beta[3] + x3[i]*beta[4] + x4[i]*beta[5] +
      x5[i]*beta[6] + x6[i]*beta[7]
  }
  #Priors:
  for(i in 1:7){
    beta[i] ~ dnorm(0, 0.001)
  }
  sigma ~ dunif(0,100)
  tau <- 1/(sigma*sigma)
}

init_values1 <- function(){
  list(beta = rnorm(7), sigma=runif(1))
}

params <- c("beta","sigma","deviance")


fit_lm1 <- jags(data=jagsdata1, inits=init_values1, parameters.to.save = params,
                                                   model.file = lm1_jags, n.chains = 3, 
                                                   n.iter = 10000, n.burnin = 200,
                                                   n.thin = 10, DIC=TRUE)
```

