---
title: "BDA_hw1"
author: "Michelle DePrenger-Levin"
date: "January 17, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("swirl")
library(swirl)
swirl()
```
info() to display options again
When the correlated regressor is omited the remaining one is biased. Not sure why the answer should be 1 (maybe run histogram of that alone first? what does it mean, the coefficient?)
Shapiro-Wilk test with normality as null hypothesis shapiro.text(fit$residuals) so significant p would mean significantly different that normal. P>0.05 'fails to reject normality'    

```{r}
#the dice question?
hist(rnorm(mean=3.5,sd = 3.5*.68,n=10))
trunc(rnorm(mean=3.5,n=10))

dice <- sample(1:6, 100, replace=TRUE)

hist(sample(1:6, 1000, replace=TRUE))
hist(dice)
head(dice,30)
```

Density plot
```{r}
# Zhang et al 2011, PlosOne
#standard deviation from standard error
n <- 80
SEmux <-0.03
#SEmux <- samplesd/sqrt(n)
samplesd <- SEmux*sqrt(n)

x<-seq(18,27,length=1000)
plot(x, dnorm(x, mean=25, sd=samplesd))
points(x, dnorm(x, mean=19, sd=samplesd),col="red")

n2<-79
SEmux2 <- 4
SEmux3 <- 3
samplesd2 <- SEmux2*sqrt(n)
samplesd3 <- SEmux3*sqrt(n)
x<-seq(86,96,length=1000)
plot(x,dnorm(x,mean=86, sd=samplesd2))
points(x, dnorm(x, mean=19, sd=samplesd2),col="red")
```

